---
title: "Lab 2 - Data wrangling"
author: "Ayden Frost"
date: "February 12, 2026"
format: pdf
---

```{r}
#| eval: true
#| message: false

library(tidyverse)
```

# Questions

## Part 1

### Question 1

```{r}
midwest |>
  count(state, sort = TRUE)
```

Illinois has the highest number of counties with 102, while Wisconsin has the lowest number of counties with only 72.

\newpage

### Question 2

```{r}
midwest |>
  count(county, state) |>
  count(county, name = "nstates") |>
  filter(nstates == 5)
```

\newpage

### Question 3

```{r}
midwest |>
  filter(popdensity > 25000) |>
  select(county, state, popdensity, poptotal, area) |>
  arrange(desc(popdensity))
```

```{r}
midwest |>
  filter(popdensity == max(popdensity)) |>
  select(county, state, popdensity, poptotal, area)
```

\newpage

### Question 4

```{r}
midwest |>
  summarize(
    median(popdensity),
    q1 = quantile(popdensity, 0.25),
    q3 = quantile(popdensity, 0.75)
  )
```

The distribution of population density of counties is unimodal and extremely right-skewed. A typical Midwestern county has population density of 1156 people per unit area. The middle 50% of the counties have population densities between 622 to 2330 people per unit area.

\newpage

### Question 5

```{r}
midwest |>
  count(state, inmetro) |>
  group_by(state) |>
  mutate(prop = n / sum(n))
```

\newpage

### Question 6

\newpage

### Question 7

\newpage

### Question 8

\newpage

## Part 2

### Question 9
